---
title: "project3_cleaning"
output:
  pdf_document: default
  html_document: default
date: "2023-03-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries

```{r}
library(tidyverse)
library(forcats)
library(plyr)
```

### Import data

```{r}
df <- read.csv('./data/kaggle_survey_2022_responses.csv',
               na.strings = c("", "NA"))
```

```{r}
#clear question text from first row
df <- df[-1,]
```

There were too many income categories for thoughtful analysis--we mapped them onto smaller groups.

```{r}
df$Q29_grouped <- mapvalues(df$Q29, 
          from=c('$0-999','1,000-1,999','2,000-2,999','3,000-3,999','4,000-4,999','5,000-7,499',
                 '7,500-9,999','10,000-14,999','15,000-19,999','20,000-24,999','25,000-29,999',
                 '30,000-39,999','40,000-49,999','50,000-59,999','60,000-69,999','70,000-79,999',
                 '80,000-89,999','90,000-99,999','100,000-124,999','125,000-149,999','150,000-199,999',
                 '200,000-249,999','250,000-299,999','300,000-499,999','$500,000-999,999','>$1,000,000'), 
          to=c('$0-4,999','$0-4,999','$0-4,999','$0-4,999','$0-4,999','$5,000-24,999','$5,000-24,999',
               '$5,000-24,999','$5,000-24,999','$5,000-24,999','$25,000-69,999','$25,000-69,999',
               '$25,000-69,999','$25,000-69,999','$25,000-69,999','$70,000-$149,999','$70,000-$149,999',
               '$70,000-$149,999','$70,000-$149,999','$70,000-$149,999','$150,000-$999,999',
               '$150,000-$999,999','$150,000-$999,999','$150,000-$999,999','$150,000-$999,999',
               '>$1,000,000')
          )
```

Define a function for organizing and plotting data for a questions which spans multiple columns.

```{r}
graph_multicolumn_q <- function(og_df, q_list, 
                                q_name, val_label,
                                q_text) {
  #Create limited df with only that question's columns
  q_df <- og_df %>%
     select(all_of(q_list))
  
  #Create long-pivoted df for graphing
  q_df1 <- q_df %>%
    pivot_longer(
    cols = everything(),
    names_to = q_name,
    values_to = val_label,
    values_drop_na = TRUE)
  
  #Produce df of value counts
  q_df1_count <- q_df1 %>%
    count(sym(val_label))
  
  #Graph result
  q_df1 %>%
    ggplot(aes(x=fct_rev(fct_infreq(!!sym(val_label)))))+
    geom_bar()+
    coord_flip()+
    theme_minimal()+
    ggtitle(q_text)+
    xlab(val_label)
  
}
```

Define a function for cross-analyzing a multi-column question against an additional 
single-column criteria (such as income, experience level or job title.)

```{r}

graph_cross_analysis <- function(og_df, q_demo, q_list, q_name, 
                                  val_label, q_text, level_order) {
  q_df <- og_df %>%
    select(all_of(c(q_demo, q_list)))
  
  q_df1 <- q_df %>%
    pivot_longer(
      cols= -1,
      names_to=q_name,
      values_to=val_label,
      values_drop_na=TRUE
  )
  
  q_df2 <- q_df1[complete.cases(q_df1), ]
  
  q_df2[[q_demo]] <- factor(q_df2[[q_demo]], levels = level_order)
  
  q_df2 %>%
    ggplot(aes(x=fct_rev(fct_infreq(!!sym(val_label))), fill = as.factor(!!sym(q_demo))))+
    facet_wrap(q_demo)+
    coord_flip()+
    geom_bar()+
    theme(axis.text.x = element_text(size = 7))+
    theme(axis.text.y = element_text(size = 7))+
    theme(legend.position = "none")+
    labs(y= "Count", x = val_label)+
    ggtitle(q_text)
    
}

```

Define column lists for each question.

```{r}

q31_cols <-  c('Q31_1','Q31_2','Q31_3','Q31_4',
               'Q31_5','Q31_6','Q31_7','Q31_8')

q32_cols <-  c('Q32')

q33_cols <-  c('Q33_1','Q33_2','Q33_3','Q33_4','Q33_5')

q34_cols <-  c('Q34_1','Q34_2','Q34_3','Q34_4','Q34_5',
               'Q34_6','Q34_7','Q34_8')

q35_cols <-  c('Q35_1','Q35_2','Q35_3','Q35_4','Q35_5','Q35_6',
               'Q35_7','Q35_8','Q35_9','Q35_10','Q35_11','Q35_12',
               'Q35_13','Q35_14','Q35_15','Q35_16')

```

Define vectors for ordering factors (later allows for graphs to be plotted properly).

```{r}
#define vectors for ordering factors

Q11_order <- c('< 1 years', '1-3 years', '3-5 years', 
  '5-10 years','10-20 years','20+ years')

Q23_order <- c('Data Scientist',
       'Data Analyst (Business, Marketing, Financial, Quantitative, etc)',
       'Currently not employed', 'Software Engineer', 'Teacher / professor',
       'Manager (Program, Project, Operations, Executive-level, etc)', 'Other',
       'Research Scientist', 'Machine Learning/ MLops Engineer',
       'Engineer (non-software)', 'Data Engineer', 'Statistician',
       'Data Architect', 'Data Administrator', 'Developer Advocate')

Q29_order <- c('$0-4,999','$5,000-24,999','$25,000-69,999','$70,000-$149,999',
               '$150,000-$999,999','>$1,000,000')

```


TESTING FACTORED GRAPH

```{r}

graph_cross_analysis2 <- function(og_df, q_demo, q_list, q_name, val_label, q_text) {
  q_df <- og_df %>%
    select(all_of(c(q_demo, q_list)))
  
  q_df1 <- q_df %>%
    pivot_longer(
      cols= -1,
      names_to=q_name,
      values_to=val_label,
      values_drop_na=TRUE
  )
  
  q_df2 <- q_df1[complete.cases(q_df1), ]
  
  q_df2 %>%
    ggplot(aes(x=fct_rev(fct_infreq(!!sym(val_label))), fill = as.factor(!!sym(q_demo))))+
    facet_wrap(q_demo)+
    coord_flip()+
    geom_bar()+
    theme(axis.text.x = element_text(size = 7))+
    theme(axis.text.y = element_text(size = 7))+
    theme(legend.position = "none")+
    labs(y= "Count", x = val_label)+
    ggtitle(q_text)
    
    
}

```

##QUESTION 31: Cloud Computing Platform Usage

```{r}
graph_multicolumn_q(df,
                    q31_cols,
                    'Question 31',
                    'Platform',
                    'Cloud Computing Platform Usage')
```


```{r}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q11', 
                     q_list = q31_cols, 
                     q_name = 'Question 31', 
                     val_label = 'Platform',
                     q_text = 'Cloud Computing Platform Usage',
                     level_order = Q11_order)
```


```{r}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q23', 
                     q_list = q31_cols, 
                     q_name = 'Question 31', 
                     val_label = 'Platform',
                     q_text = 'Cloud Computing Platform Usage',
                     level_order = Q23_order)
```


```{r}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q29_grouped', 
                     q_list = q31_cols, 
                     q_name = 'Question 31', 
                     val_label = 'Platform',
                     q_text = 'Cloud Computing Platform Usage',
                     level_order = Q29_order)
```


##QUESTION 32: Best Developer Experience

```{r}
graph_multicolumn_q(df,
                    q32_cols,
                    'Question 32',
                    'Platform',
                    'Cloud Platform: Best Dev. Experience')
```


###Experience
```{r}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q11', 
                     q_list = q32_cols, 
                     q_name = 'Question 32', 
                     val_label = 'Platform',
                     q_text = 'Cloud Platform: Best Dev. Experience',
                     level_order = Q11_order)
```

###Job Title
```{r}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q23', 
                     q_list = q32_cols, 
                     q_name = 'Question 32', 
                     val_label = 'Platform',
                     q_text = 'Cloud Platform: Best Dev. Experience',
                     level_order = Q23_order)
```

###Income
```{r}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q29_grouped', 
                     q_list = q32_cols, 
                     q_name = 'Question 32', 
                     val_label = 'Platform',
                     q_text = 'Cloud Platform: Best Dev. Experience',
                     level_order = Q29_order)
```


##QUESTION 33: Cloud Computing Product Usage

```{r}
graph_multicolumn_q(df,
                    q33_cols,
                    'Question 33',
                    'Platform',
                    'Cloud Computing Product Usage')
```


###Experience
```{r}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q11', 
                     q_list = q33_cols, 
                     q_name = 'Question 33', 
                     val_label = 'Platform',
                     q_text = 'Cloud Computing Product Usage',
                     level_order = Q11_order)
```

###Job Title
```{r}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q23', 
                     q_list = q33_cols, 
                     q_name = 'Question 33', 
                     val_label = 'Platform',
                     q_text = 'Cloud Computing Product Usage',
                     level_order = Q23_order)
```

###Income
```{r}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q29_grouped', 
                     q_list = q33_cols, 
                     q_name = 'Question 33', 
                     val_label = 'Platform',
                     q_text = 'Cloud Computing Product Usage',
                     level_order = Q29_order)
```

##QUESTION 34: Data Storage Product Usage

```{r}
graph_multicolumn_q(df,
                    q34_cols,
                    'Question 34',
                    'Platform',
                    'Data Storage Product Usage')
```

###Experience
```{r}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q11', 
                     q_list = q34_cols, 
                     q_name = 'Question 34', 
                     val_label = 'Platform',
                     q_text = 'Data Storage Product Usage',
                     level_order = Q11_order)
```


###Job Title
```{r}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q23', 
                     q_list = q34_cols, 
                     q_name = 'Question 34', 
                     val_label = 'Platform',
                     q_text = 'Data Storage Product Usage',
                     level_order = Q23_order)
```


###Income
```{r}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q29_grouped', 
                     q_list = q34_cols, 
                     q_name = 'Question 34', 
                     val_label = 'Platform',
                     q_text = 'Data Storage Product Usage',
                     level_order = Q29_order)
```


##QUESTION 35: Additional Data Products (databases, data warehouses, data lakes)

```{r}
graph_multicolumn_q(df,
                    q35_cols,
                    'Question 35',
                    'Platform',
                    'Databases, Warehouses and Lakes')
```

###Experience
```{r}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q11', 
                     q_list = q35_cols, 
                     q_name = 'Question 35', 
                     val_label = 'Platform',
                     q_text = 'Databases, Warehouses and Lakes',
                     level_order = Q11_order)
```


###Job Title
```{r, fig.width=10, fig.height=10}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q23', 
                     q_list = q35_cols, 
                     q_name = 'Question 35', 
                     val_label = 'Platform',
                     q_text = 'Databases, Warehouses and Lakes',
                     level_order = Q23_order)
```

###Income
```{r}
graph_cross_analysis(og_df = df, 
                     q_demo = 'Q29_grouped', 
                     q_list = q35_cols, 
                     q_name = 'Question 35', 
                     val_label = 'Platform',
                     q_text = 'Databases, Warehouses and Lakes',
                     level_order = Q29_order)
```


